---
title: "Introduction to Smoothing"
subtitle: "Machine Learning - Section $3.2.1$"
author: "_Marc Omar Haddad_"
date: "Published: 25 February, 2020 <br>"
output:
  html_notebook:
    theme: readable
    highlight: zenburn
  # pdf_document:
  #   latex_engine: xelatex
  #   df_print: kable
---
<!-- FOR UPDATES -->
  
  <!-- Uncomment if output = html_notebook: -->
     <center><strong>Updated: `r format(Sys.time(), '%d %B, %Y')`</strong></center>
  
  <!-- Uncomment if output = pdf_document: -->
    <!-- \begin{center}Updated: `r format(Sys.time(), '%d %B, %Y')`\end{center} -->

<!-- FOR UPDATES -->    

\

```{r dependencies, include=FALSE, results='hide'}
library(tidyverse)
library(caret)
library(dslabs)
library(ggrepel)
```

\

Smoothing is an extremely powerful tool that is used accross data analysis fields. Also known as "Curve Fitting" and "Low Band Pass Filtering", **Smoothing** is designed to **detect trends** in the presence of **noisy data** (where the shape of the trend is unknown). The term "Smoothing" is used to indicate that **we assume the trend of noisy data to be smooth**. 

Smoothing techniques are quite useful when applied in a machine learning context. In machine learning, like in Smoothing, our goal is to **extract conditional expectations/probabilities** (the underlying trends) of an unknown shape, in the presence of **uncertainty** (noisy data).


### Introduction to our Example Problem: Popularity Over Time

We will be trying to estimate the **time trend of the popularity vote** in the 2008 Presidential election between Obama and McCain.

```{r load_data_and_plot, echo=TRUE}
data("polls_2008")
qplot(day, margin, data = polls_2008)
```



Note: **Do not think of this example as a forecasting problem**. All we are interested in is **learning the shape of the trend _after_ collecting all the data**.

\

We assume that **for every day $x$,** we have a **true preference among the electorate, $f(x)$**. However, due to the uncertainty of polling, each data point comes with an **error $\epsilon$.**  

The following is the mathematical model for the observed poll margin $Y$:

$$
Y_i = f(x_i) + \epsilon_i
$$

We can redefine this problem in a machine learning context: We want to **predict $Y$ given day $x$, through the conditional expectation of**:
$$
f(x) = \mathrm{E}(Y\ |\ X = x)
$$
Since we *don't know* the true value of the above, we must **estimate it**. 

We will first use regression for our estimates. 

```{r reg_attempt, echo=FALSE}
fit = lm(polls_2008$margin ~ polls_2008$day)
qplot(day, margin, data = polls_2008, color=ifelse(fit$residuals > fit$coefficients[2], "+", "-")) +
geom_abline(aes(intercept=fit$coefficients[1],slope=fit$coefficients[2])) + theme(legend.title=element_blank())

```


As we can see in our above plot, our regression line does not do a great job of describing our trend. We will need a more flexible algorithm.

